---
title: "Coursera Assignment"
author: "Gmatmarshall"
date: "September 11, 2016"
output: html_document
---

## Loading and preprocessing the data
The following command uses R to download and unzip the data file and store the results in the variable activity.

```{r, echo=TRUE}
setwd("~/Coursera/Reproducible_Research")
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip","activity.zip",method="curl")
unzip("activity.zip", "activity.csv")
activity <- read.csv("~/Coursera/Reproducible_Research/activity.csv")
```

The next command reoves the NA elements from the dataset and store the new data in the variable activityn
```{r, echo=TRUE}
activityn<-activity[!is.na(activity$steps),]
```

## The mean total number of steps taken per day
Now the total number of steps each day is computed and stored in the variable sumd.
```{r, echo=TRUE}
sumd<-with(activityn,tapply(steps,date,sum))
```

```{r, echo=TRUE}
hist(sumd, main="Histogram of oringial data",xlab="steps")
```

The mean and median of the total number of steps taken per day can easily be computed.
```{r, echo=TRUE}
mean(sumd,na.rm=TRUE)
median(sumd,na.rm= TRUE)
```

## The average daily activity pattern

Using activityn variable (where the NAs have been removed) we compute the average across all days for particular
intervals.

```{r, echo=TRUE}
avgd<- with(activityn,tapply(steps,interval,mean))
plot(avgd~as.integer(names(avgd)),type='l', main = "Time series plot of steps", xlab="intervals")

```

The interval that contains the maximum is 
```{r, echo=TRUE}
avgd[avgd==max(avgd)]
```
We see that interval 835 contains the highest number of steps.

## Imputing missing values

The code below will count the number of NAs (done by coercing the TRUE values to 1 and FALSE values to 0)
```{r, echo=TRUE}
sum(is.na(activity$steps))
```


Replace all the NA values with the mean and store it in variable activity2
```{r, echo=TRUE}
activity2<-activity
activity2$steps[is.na(activity$steps)]<-mean(activity$steps,na.rm=TRUE)
```

Note the histogram is the same, which means it might be possible that base R is handling the missing values for histogram by imputing the mean.

```{r, echo=TRUE}
sumd2<-with(activity2,tapply(steps,date,sum))
hist(sumd2, main="Histogram of data with filled in missing values", xlab="interval")
mean(sumd2)
median(sumd2)
```

Note these values do not change much from the original values. It will affect the standard deviation.

## Differences in activity patterns between weekdays and weekends?

The code below converts the dates in to the form of factors (weekdays and weekends)
```{r,echo=TRUE}
activity3<- activity2
 activity3$date<- as.Date(activity3$date)
 activity3$date<-weekdays(activity3$date)
activity3$date<-as.factor(activity3$date)
activity4<-activity3
levels(activity4$date)
levels(activity4$date) <- c("weekday","weekday","weekend","weekend","weekday","weekday","weekday")
```

```{r,echo=TRUE}
wkday<-activity4[activity4$date=="weekday",]
wkend<-activity4[activity4$date=="weekend",]

meanwkday<-with(wkday,tapply(steps,interval,mean))
meanwkend<-with(wkend,tapply(steps,interval,mean))

par(mfrow=c(2,1))
plot(meanwkend~as.integer(names(meanwkend)),type='l', main="Weekend Steps",xlab="steps")
plot(meanwkday,type='l', main="Weekday Steps", xlab="steps",ylab="average")
```

We note that there is more activity on the weekend after the 1000th interval.